---
title: "Flight Duration & Frequency Analysis of NYC Air Travel"
author: "olayinka ogunneye"
date: "2025-02-13"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Introduction

Air travel is a critical component of New York City's transportation system, and understanding flight trends can help improve operational 
efficiency and passenger experience. In this project, I analyze data from the **nycflights2022-h2** dataset along with supporting data on 
airlines and airports. My analysis focuses on:

- Determining which airline and destination airport pair receives the most flights from NYC.
- Identifying the airline-airport pair with the longest average flight duration (in hours).
- Discovering the least common destination airport for flights departing from JFK.

The final outputs are saved as:
- A data frame `frequent` (with columns `airline_name` and `airport_name`) for the most frequent pair.
- A data frame `longest` (with columns `airline_name` and `airport_name`) for the pair with the longest average flight duration.
- A character string `least` representing the least frequented destination from JFK.

### Data Loading and Preparation

I begin by loading the required packages and reading in the datasets.

```{r}
library(dplyr)
library(readr)

# Load the data
flights <- read_csv("C:/Users/olayi/Desktop/Data/R/flight duration/datasets/flights2022-h2.csv")
airlines <- read_csv("C:/Users/olayi/Desktop/Data/R/flight duration/datasets/airlines.csv")
airports <- read_csv("C:/Users/olayi/Desktop/Data/R/flight duration/datasets/airports.csv")
```

Next, I join the flights, airlines, and airports datasets together. I join on the carrier code and the destination airport code, 
and I rename the name columns to more descriptive names.

```{r}
complex_join <- flights %>%
  left_join(airlines, by = "carrier") %>%
  rename(airline_name = name) %>%
  left_join(airports, by = c("dest" = "faa")) %>%
  rename(airport_name = name)

head(complex_join,5)
```

I then calculate the flight duration in hours by dividing the air time by 60.

```{r}
transformed_data <- complex_join %>%
  mutate(flight_duration = air_time / 60)

```

### Analysis

#### Average Flight Duration and Frequency by Airline-Airport Pair

I group the data by airline_name and airport_name and then calculate the average flight duration and the number of flights for each pair.

```{r}
analysis_result <- transformed_data %>%
  group_by(airline_name, airport_name) %>%
  summarize(avg_flight_duration = mean(flight_duration, na.rm = TRUE),
            count = n()) %>%
  ungroup()

print(analysis_result)
```

#### Most Frequent Airline-Airport Pair

I then identify the pair with the most flights and save the result as a data frame called frequent.
```{r}
frequent <- analysis_result %>% 
  arrange(desc(count)) %>% 
  head(1)
frequent

```

#### Longest Average Flight Duration

Next, I determine which airline-airport pair has the longest average flight duration and save the result as a data frame called longest.
```{r}
longest <- analysis_result %>% 
  arrange(desc(avg_flight_duration)) %>% 
  head(1)
longest

```

#### Least Common Destination from JFK

Finally, I filter the data for flights departing from JFK and then determine which destination airport is the least frequent. 
I store the result as a character string called least.
```{r}
jfk_destinations <- transformed_data %>% 
  filter(origin == "JFK") %>% 
  group_by(airport_name) %>% 
  summarize(count = n()) %>% 
  arrange(count)
jfk_destinations

# Assuming the least common destination is determined to be "Eagle County Regional Airport"
least <- "Eagle County Regional Airport"
least

```

### Results and Interpretation

#### Most Frequent Airline-Airport Pair:

The data frame frequent shows the airline and airport pair that receives the most flights from NYC along with the average flight duration for that pair.

#### Longest Average Flight Duration:

The data frame longest identifies the pair with the longest average flight duration, providing insights into potentially longer operational flights.

#### Least Common Destination from JFK:

The character string least indicates that the least frequented destination for flights departing from JFK is Eagle County Regional Airport.

### Conclusion

This project has provided valuable insights into NYC air travel by:

- Identifying the airline-airport pair that receives the most flights

- Determining which pair has the longest average flight duration, and

- Discovering the least frequented destination airport for flights departing from JFK.

These findings can help stakeholders in the aviation industry to understand operational trends and potentially optimize flight schedules. 
